### 2004-07-05 Documentation/part1.tex
Prior handling updated

@@

@@

Before we can set the simulated event rates or access any oscillation
parameters, we need to become familiar with the concept \GLOBES\ uses
for oscillation parameters. In order to transfer sets of oscillation
parameter vectors $(\theta_{12}, \theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm)$, the parameter type \GLB{glb\_params\_type} is used. In general, pointers to this structure are of the type \GLB{glb\_params}, and are often
transferred to and from \GLOBES\ functions. Therefore, the memory space
for these vectors has to be reserved before they can be used, and it has
to be returned afterwards. \GLOBES\ functions usually use the pointers
of the type {\tt glb\_params} for the input or output to the functions. As
an input parameter, the pointer has to be created and point towards a valid
parameter structure, where the oscillation parameters are read from. As
an output parameter, the pointer has to be created, too, and point towards
a structure which will contain the return values will be written to.
This parameter transfer concept seems to be very sophisticated, but, as we
will see in the next chapters, it hides a lot of complicated parameter 
mappings which otherwise need to be done by the user. For example, not
only the oscillation parameters are stored in the pointer structure, but also information on the
densities of all of the initialized experiments. Since \GLOBES\ treats
the matter density as a free parameter known with some external precision
to include matter density uncertainties,
the minimizers also use fit values and external errors for the matter
densities of all loaded experiments. More precisely, the matter density
profile of each experiment $i$ is multiplied by a scaling factor
$\hat{\rho_i}$, \index{Density scaling factor} 
which is stored in the density information of {\tt glb\_params}. Each of
these scaling factors has $1.0$ as pre-defined value. Since it
is in most cases not necessary to change this value, the user does
not need to take care of these values.\footnote{For more information,
we refer to the following chapters.}
Note that the {\tt glb\_params}
structures must not be initialized before all experiments are loaded,
since the number of matter densities can only be determined after the
experiments are initialized. 
Similarly, any change in the number of experiments requires that the
parameter structures be re-initialized, \ie, destroyed and created again.
Another piece of information returned from the minimizers and transferred
into the {\tt glb\_params} structure is the number
of iterations used for the minimization, which is proportional to the
running time of the minimizer. In general, the user does not need to
access the elements in {\tt glb\_params} directly. A number of functions
is provided to handle these parameter structures:
\GLBNS{glbAllocParams}
{\tt glb\_params glbAllocParams()} allocates the memory space
needed for a parameter vector and returns a pointer to it.
\GLBNS{glbAllocParams}
{\tt void glbFreeParams(glb\_params stale)} destroys the memory space
needed for a parameter vector {\tt stale} and sets the pointer to {\tt NULL}.
\GLBNS{glbDefineParams}
{\tt glb\_params glbDefineParams(glb\_params in, double theta12, double theta13,double theta23, double delta, double dms, double dma)} assigns
the complete set of oscillation parameters to the vector {\tt in}, which has
to be allocated before. The return value is the pointer to {\tt in} if 
the assignment was successful, and {\tt NULL} otherwise.
 \GLBNS{glbCopyParams}
 {\tt glb\_params glbCopyParams(const glb\_params source, glb\_params dest)}
 copies the vector {\tt source} to the vector {\tt destination}. The return
 value is {\tt NULL} if the assignment was not successful.
\GLBNS{glbPrintParams}
{\tt void glbPrintParams(FILE *stream, const glb\_params in)} prints the
parameters in {\tt in} to the file {\tt stream}. The oscillation
parameters, all density values, and the number of iterations are
printed as pretty output. Use {\tt stdout} for {\tt stream} if you want
to print to the screen.
In addition to these basic functions, there are functions to access
the individual parameters within the parameter vectors:
\GLBNS{glbSetOscParams}
{\tt glb\_params glbSetOscParams(glb\_params in, double osc, int which)}
sets the oscillation parameter {\tt which} in the structure {\tt in}
to the value {\tt osc}. If the assignment was unsuccessful, the 
function returns {\tt NULL}.
\GLBNS{glbGetOscParams}
{\tt double glbGetOscParams(glb\_params in, int which)}
returns the value of the oscillation parameter {\tt which} in the 
structure {\tt in}.
In both of these functions, the parameter {\tt which} runs from
$0$ to $5$, where the parameters in \GLOBES\ always have the order $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, $\deltacp$, $\sdm$, $\ldm$.
Alternatively to the number, the constants \GLB{GLB\_THETA\_12}, \GLB{GLB\_THETA\_13}, \GLB{GLB\_THETA\_23}, \GLB{GLB\_DELTA\_CP}, 
\GLB{GLB\_DM\_SOL}, or \GLB{GLB\_DM\_ATM} can be used.

Similarly, the density parameters or iterations can be accessed:
\begin{function}
\GLBNS{glbSetDensityParams}
{\tt glb\_params glbSetDensityParams(glb\_params in, double dens, int which)} sets the density parameter {\tt which} in the structure {\tt in}
to the value {\tt dens}. If the assignment was unsuccessful, the 
function returns {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{glbGetDensityParams}
{\tt double glbGetDensityParams(glb\_params in, int which)}
returns the value of the density parameter {\tt which} in the 
structure {\tt in}.
\end{function}
\begin{function}
\GLBNS{glbSetIteration}
{\tt glb\_params glbSetIteration(glb\_params in, int iter)} sets the number of iterations in the structure {\tt in}
to the value {\tt iter}. If the assignment was unsuccessful, the 
function returns {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{glbGetIteration}
{\tt int glbGetIteration(glb\_params in)}
returns the value of the number of iterations in the 
structure {\tt in}.
\end{function}
In total, the parameter vector handling in a program normally has the
following order:
\begin{quote}
{\tt
glbInitExperiment(...); \\
/* ... more initializations ... */ \\
\\
glb\_params vector1 = glbAllocParams(); \\
/* ... more vectors allocated ... */ \\
\\
/* Program code: assign and use vectors */ \\
\\
glbFreeParams(vector1); \\
/* ... more vectors destroyed ... */ \\
\\
/* ... end of program or glbClearExperimentList ... */ 
}
\end{quote}
Compared to existing experiments, which use real data, future experiments
uses simulated data. Thus, the {\em true parameter values} and their results in form of the reference event rate vectors are simulated. After setting the true parameter values, the {\em fit parameter values} can be varied in order to obtain information on the measurement performance for the given set of true parameter values. Therefore, it is often useful to show the results of a future measurement as function of the true parameter values for which the reference rate vectors are computed -- at least within the currently allowed ranges. The true parameter values for the vacuum neutrino oscillation parameters have to be set by the functions \GLB{glbSetOscillationParameters} and \GLB{glbSetRates} {\em before} any evaluation function is used and {\em after} the experiments have been initialized and the experiment parameters have been adjusted which could change the rates (such as baseline or target mass). This means that
after any change of an experiment parameter, {\tt glbSetRates} has to be called. Matter effects are automatically included as specified in the experiment definition. We have the following functions to assign and
read out the vacuum oscillation parameters:
\begin{function}
\GLBNS{glbSetOscillationParameters}
{\tt int glbSetOscillationParameters(const glb\_params in)} sets the
vacuum oscillation parameters to the ones in the vector {\tt in}.
\end{function}
\begin{function}
\GLBNS{glbGetOscillationParameters}
{\tt int glbGetOscillationParameters(glb\_params out)} returns the
vacuum oscillation parameters in the vector {\tt out}. The result of
the function is $0$ if the call was successful.
\end{function}
The reference rate vector is then computed with:
\begin{function}
\GLBNS{glbSetRates}
{\tt void glbSetRates()} computes the reference rate vector for the neutrino oscillation parameters set by {\tt glbSetOscillationParameters}. 
\end{function}
A complete example for a minimal \GLOBES\ program can be found on
Page~\pageref{ex:c}.

@@

\index{Systematics}
Calculating a $\chi^2$-value with or without systematics, but no correlations and degeneracies, is the simplest and fastest possibility to obtain high-level information on an experiment. I general, \GLOBES\ uses the six independent oscillation parameters $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, $\deltacp$, $\sdm$, $\ldm$, as well as the matter density of each experiment. Thus, there are six plus the number of experiments parameters determining the rate vectors. Using the matter densities in addition to the oscillation parameters will allow the simulation of matter density uncertainties: In this approach, the matter densities can be treated as parameters to be measured by the experiments within certain limits, where the limits are given by the observed precision of the matter density profile. Defining a independent matter density for each experiment is necessary if the baselines are completely uncorrelated. For correlated or even identical baselines, one may rather want to encapsulate the experiments into a single experiment definition.

\example{Correlation between $\stheta$ and $\deltacp$}{
\label{ex:corrth13dcp}
\index{Two-parameter correlation}
A typical and fast application for \GLB{glbChiSys} is the visualization of
 two-parameter correlations. For example, to calculate the two-parameter 
 correlation between $\stheta$ and $\deltacp$ at a neutrino factory, one
 can use the following code excerpt from {\tt example1.c}:
\begin{quote}
{\tt {\footnotesize
  /* Initialize parameter vector(s) and compute simulated data */ \\
  glbDefineParams(true\_values,theta12,theta13,theta23,deltacp,sdm,ldm); \\
  glbDefineParams(test\_values,theta12,theta13,theta23,deltacp,sdm,ldm);  \\
  glbSetOscillationParameters(true\_values); glbSetRates(); \\
\\
  /* Iteration over all values to be computed */ \\
  double x,y,res; \\   
  for(x=-4.0;x<-2.0+0.01;x=x+2.0/50) \\
  for(y=0.0;y<200.0+0.01;y=y+200.0/50) \\
  \{ \\
\hspace*{0.5cm} /* Set parameters in vector of test values */ \\
\hspace*{0.5cm} glbSetOscParams(test\_values,asin(sqrt(pow(10,x)))/2,GLB\_THETA\_13); \\
\hspace*{0.5cm} glbSetOscParams(test\_values,y*M\_PI/180.0,GLB\_DELTA\_CP); \\
  \\
\hspace*{0.5cm} /* Compute Chi2 for all loaded experiments and all rules */ \\
\hspace*{0.5cm} res=glbChiSys(test\_values,GLB\_ALL,GLB\_ALL); \\
\hspace*{0.5cm} AddToOutput(x,y,res); \\
  \} \\
} }
\end{quote}

@@

Keeping all oscillation parameters and matter densities fixed, one can use the following functions to obtain the total $\chi^2$ of all specified oscillation channels including systematics:
\begin{function} 
\GLB{Chi}$(\{ \theta_{12}, \theta_{13}, \theta_{23}, \deltacp , \sdm , \ldm, \hat{\rho_1}, \hdots , \hat{\rho_n} \})$  returns the total added $\chi^2$ of all loaded experiments.
\end{function}
\begin{function}
 \GLB{SingleChi}$(\{ \theta_{12}, \theta_{13}, \theta_{23}, \deltacp , \sdm , \ldm, \hat{\rho}_{N_{\mathrm{exp}}} \}, \, N_{\mathrm{exp}} )$  returns the total $\chi^2$ of the experiment number $N_{\mathrm{exp}}$.
\end{function}
Note that the result of {\tt Chi} corresponds to the sum of all of the {\tt SingleChi}'s of the loaded experiment. This equality will not hold for the minimizors in the next sections anymore. In both functions, one has to give the matter density scaling factor $\hat{\rho_i}$ for each of the used experiments. The effect of this scaling factor depends on the type of the matter density profile, which is given in the  experiment definition. For a constant matter density, it is simply the ratio of the matter density and the average matter density specified in the experiment definition, \ie , $\hat{\rho_i} \equiv \rho_i/\bar{\rho}_i$.\index{Matter scaling factor} For a matter density profile, it is an overall scaling factor: The matter density in each layer is multiplied by this factor. In most cases one wants to take a scaling factor of $1.0$ here, which simply means taking the matter density profile as it is given in the experiment definition. Moreover, note that its effect is in general small for short baselines. An example of how to use  {\tt Chi} (or {\tt SingleChi}) can be found on page~\pageref{ex:corrth13dcp}.

\index{Systematics}
The treatment of systematics is done by the usage of auxiliary systematics parameters, which are taken completely uncorrelated among different oscillation channels, and treated with simple Gaussian statistics. One such example is the signal normalization error, \ie, an error to the overall normalization of the signal. For illustration, we assume that the signal event rate in the $i$th bin $s_i^0$ of one oscillation channel is altered by the overall normalization auxiliary parameter\index{Auxiliary parameter} of this channel, \ie , 
\be
 s_i = s_i(n_s) = s_i^0 \cdot (1 + n_s),
\ee
where $n_s$ is the signal normalization parameter. The total number of events in the $i$th bin $x_i$ also includes the background event rates $b_i$, \ie, $x_i = s_i + b_i$, which may have their own systematics parameters.
In order to implement an overall signal normalization error $\sigma_{n_s}$,  the $\chi^2$, which includes all event rates $x_i$ of all bins, is minimized over the auxiliary parameter $n_s$:
\be
 \hat{\chi^2} = \underset{n_s}{\mathrm{min}} \left(  \chi^2(n_s, \hdots) + \frac{(n_s)^2}{\sigma_{n_s}^2} \right).
\ee 
This minimization is done independently for all auxiliary parameters of this oscillation channel. The total $\chi^2$ for the considered experiment is finally obtained by repreating this procedure for all oscillation channels and adding their $\chi^2$-values. In general, the situation is more complicated because of the usage of many systematical errors. More details about systematics parameters and the definition of signal, background, and oscillation channels can be found in the \EDM\ part of this book, too.

The systematics minimization of an experiment can be easily switched on and off, \ie, one can also compute the $\chi^2$ without even taking into account systematics:
\begin{function}
\GLB{SetSystematics}$(N_{mathrm{exp}}, \, sys)$ switches the systematics minimization for experiment $N_{\mathrm{exp}}$ on (sys=1) or off (sys=0).
\end{function}
This function can be especially useful for the test of the impact of systematics.
??? Wie einzelne Systematik-Parameter ein und ausschalten? Sonst individuell nicht testbar! ??? Verweis auf appendix?

\chapter[Calculating $\chi^2$-projections: how one can include correlations]{Calculating $\boldsymbol{\chi^2}$-projections: how one can include correlations}

\index{Multi-parameter correlation}
This chapter deals with the rather complicated issue of $n$-parameter correlations. Since it has before this software  not been possible to include the full $n$-parameter correlations in the high-dimensional parameter space with reasonable effort, it is the core part of this software  -- as well as its strength. Of course, calculating $\chi^2$-projections is somewhat more complicated than using systematics only. Therefore, we use a simple step by step introduction to the problem.

@@

\index{Projection of manifold}
In principle, the precision of an individual parameter measurement including correlations can be obtained as the projection of the $n$-dimensional fit manifold onto the respective axis. Similarly, one can project the fit manifold onto a plane, such as the $\stheta$-$\deltacp$-plane, if one wants to explicitely show this correlation with all the other parameter correlations included. In practice, this projection is very difficult: a grid-based method would need $(N_{\mathrm{grid}})^n$ function calls of {\tt Chi} or {\tt SingleChi} to calculate the precision including the full $n$-parameter correlation, where $N_{\mathrm{grid}}$ is the number of points in each direction of the lattice. For example, taking only $N_{\mathrm{grid}}=20$ and $n=7$ (six oscillation parameters and matter density) would mean more than one billion function calls of {\tt Chi} or {\tt SingleChi}. One can easily image that this is too much for any sophisticated application.

\index{Minimizer}
The solution to this problem is using a local $n$-dimensional minimizer instead of a grid-based method. It turns out that such a minimizer can include a full $6$-parameter correlation with of the order of $1\, 000$ function calls of {\tt Chi} or {\tt ChiNew}. It is a standard method which can be found in every good book for standard numerical calculation routines. Thus, for each point on the projection axis/plane, one can obtain a result within about $10$ to $30$ seconds on a modern computer, which means that the complete measurement precision for one fixed true parameter set can be obtained in as much as $10$ to $15$ minutes. One can easily imagine that such a minimizer makes more sophisticated applications possible with the help of overnight calculations, such as showing the dependencies on the true parameter values.

This approach also has a major disadvantage: One can not simply program a robust grid-based code and let it run, since using a local minimizer always means that one may end up in an unwanted local minimum and not in the investigated one. Thus, one has to use some (analytical or numerical) knowledge on the topology of the fit manifold and start the local minimizer close enough to the investigated solution. Fortunately, this can be done quite straightfoward in most cases, since the structure of the neutrino oscillation formulas does not cause very complicated topologies of the fit manifolds. Especially, the are plenty of analytical discussions of this issue, which means that one can implicitely use this knowledge to obtain better predictions for the measurement performances. Note that, since one can easily find the global fit minimum at the best-fit values, any solution found with the local minimizer makes the measurement performance worse. Thus, one can only run the danger to obtain a too optimistic solution if one does not find the other local minima below the chosen confidence level.

\index{Priors} \index{Input errors} \index{Starting values}
In many cases, the fit manifold is restricted by the knowledge from earlier experiments. For example, the knowledge on the solar parameters will in most cases be supplied by the solar neutrino experiments. If the external precision of a parameter is at the time of the measurement better than the one of the experiment itself, one has to impose some external knowledge on this parameter. This external knowledge may reduce the $n$-dimensional fit manifold in the respective direction. In the most extreme case, keeping all parameters but the measured one fixed in the analysis means that all parameters are determined externally with infinitively high precisions. Thus, using the projection method on the axis/plane of interest is a reasonable approach. The inclusion of external input in \GLOBES\ is done by the use of Gaussian {\em priors}: We assume that an external measurement has determined the measured parameter to be at the central value (called {\em starting value}) with a $1 \sigma$ Gaussian error (called {\em input error}). The explicit definition of these priors will be shown in the next section.

>>>>

\section{Initialization of \GLOBES}

@@

\section{Units in \GLOBES\ and the integrated luminosity}

@@

\section{Handling oscillation parameter vectors}

Before we can set the simulated event rates or access any oscillation
parameters, we need to become familiar with the concept \GLOBES\ uses
for oscillation parameters. In order to transfer sets of oscillation
parameter vectors $(\theta_{12}, \theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm)$, the parameter type \GLB{glb\_params\_type} is used. In general, pointers to this structure are of the type \GLB{glb\_params}, and are often
transferred to and from \GLOBES\ functions. Therefore, the memory space
for these vectors has to be reserved before they can be used, and it has
to be returned afterwards. \GLOBES\ functions usually use the pointers
of the type {\tt glb\_params} for the input or output to the functions. As
an input parameter, the pointer has to be created and point towards a valid
parameter structure, where the oscillation parameters are read from. As
an output parameter, the pointer has to be created, too, and point towards
a structure which will contain the return values will be written to.
This parameter transfer concept seems to be very sophisticated, but, as we
will see in the next chapters, it hides a lot of complicated parameter 
mappings which otherwise need to be done by the user. For example, not
only the oscillation parameters are stored in the pointer structure, but also information on the matter
densities of all of the initialized experiments. Since \GLOBES\ treats
the matter density as a free parameter known with some external precision
to include matter density uncertainties,
the minimizers also use fit values and external errors for the matter
densities of all loaded experiments. More precisely, the matter density
profile of each experiment $i$ is multiplied by a scaling factor
$\hat{\rho_i}$, \index{Matter density scaling factor} 
which is stored in the density information of {\tt glb\_params}. 
Each of
these scaling factors has $1.0$ as pre-defined value. Since it
is in most cases not necessary to change this value, the user does
not need to take care of it. For a constant matter density, it is simply the ratio of the matter density and the average matter density specified in the experiment definition, \ie , $\hat{\rho_i} \equiv \rho_i/\bar{\rho}_i$. For a matter density profile, it acts as an overall normalization factor: The matter density in each layer is multiplied by this factor. In most cases one wants to take a scaling factor of $1.0$ here, which simply means taking the matter density profile as it is given in the experiment definition. For the treatment of correlations, however,
an external precision of the scaling factor might be used to include the
correlations with the matter density uncertainty. For more details, we
refer to the following sections.
Note that the {\tt glb\_params}
structures must not be initialized before all experiments are loaded,
since the number of matter densities can only be determined after the
experiments are initialized. 
Similarly, any change in the number of experiments requires that the
parameter structures be re-initialized, \ie, destroyed and created again.

Another piece of information will be returned from the minimizers 
(\cf, \Chapt~\ref{chapt:correlations}) and transferred
into the {\tt glb\_params} structure is the number
of iterations used for the minimization, which is proportional to the
running time of the minimizer. In general, the user does not need to
access the elements in {\tt glb\_params} directly. A number of functions
is provided to handle these parameter structures:
\begin{function}
\GLBNS{glbAllocParams}
{\tt glb\_params glbAllocParams()} allocates the memory space
needed for a parameter vector and returns a pointer to it.
\end{function}
\begin{function}
\GLBNS{glbAllocParams}
{\tt void glbFreeParams(glb\_params stale)} destroys the memory space
needed for a parameter vector {\tt stale} and sets the pointer to {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{glbDefineParams}
{\tt glb\_params glbDefineParams(glb\_params in, double theta12, double theta13,double theta23, double delta, double dms, double dma)} assigns
the complete set of oscillation parameters to the vector {\tt in}, which has
to be allocated before. The return value is the pointer to {\tt in} if 
the assignment was successful, and {\tt NULL} otherwise.
\end{function}
\begin{function}
 \GLBNS{glbCopyParams}
 {\tt glb\_params glbCopyParams(const glb\_params source, glb\_params dest)}
 copies the vector {\tt source} to the vector {\tt destination}. The return
 value is {\tt NULL} if the assignment was not successful.
\end{function}
\begin{function}
\GLBNS{glbPrintParams}
{\tt void glbPrintParams(FILE *stream, const glb\_params in)} prints the
parameters in {\tt in} to the file {\tt stream}. The oscillation
parameters, all density values, and the number of iterations are
printed as pretty output. Use {\tt stdout} for {\tt stream} if you want
to print to the screen.
\end{function}
In addition to these basic functions, there are functions to access
the individual parameters within the parameter vectors:
\begin{function}
\GLBNS{glbSetOscParams}
{\tt glb\_params glbSetOscParams(glb\_params in, double osc, int which)}
sets the oscillation parameter {\tt which} in the structure {\tt in}
to the value {\tt osc}. If the assignment was unsuccessful, the 
function returns {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{glbGetOscParams}
{\tt double glbGetOscParams(glb\_params in, int which)}
returns the value of the oscillation parameter {\tt which} in the 
structure {\tt in}.
\end{function}
In both of these functions, the parameter {\tt which} runs from
$0$ to $5$, where the parameters in \GLOBES\ always have the order $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, $\deltacp$, $\sdm$, $\ldm$.
Alternatively to the number, the constants \GLB{GLB\_THETA\_12}, \GLB{GLB\_THETA\_13}, \GLB{GLB\_THETA\_23}, \GLB{GLB\_DELTA\_CP}, 
\GLB{GLB\_DM\_SOL}, or \GLB{GLB\_DM\_ATM} can be used.

Similarly, the density parameters or iterations can be accessed:
\begin{function}
\GLBNS{glbSetDensityParams}
{\tt glb\_params glbSetDensityParams(glb\_params in, double dens, int which)} sets the density parameter {\tt which} in the structure {\tt in}
to the value {\tt dens}. If the assignment was unsuccessful, the 
function returns {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{glbGetDensityParams}
{\tt double glbGetDensityParams(glb\_params in, int which)}
returns the value of the density parameter {\tt which} in the 
structure {\tt in}.
\end{function}
\begin{function}
\GLBNS{glbSetIteration}
{\tt glb\_params glbSetIteration(glb\_params in, int iter)} sets the number of iterations in the structure {\tt in}
to the value {\tt iter}. If the assignment was unsuccessful, the 
function returns {\tt NULL}.
\end{function}
\begin{function}
\GLBNS{glbGetIteration}
{\tt int glbGetIteration(glb\_params in)}
returns the value of the number of iterations in the 
structure {\tt in}.
\end{function}
In total, the parameter vector handling in a program normally has the
following order:
\begin{quote}
{\tt
glbInitExperiment(...); \\
/* ... more initializations ... */ \\
\\
glb\_params vector1 = glbAllocParams(); \\
/* ... more vectors allocated ... */ \\
\\
/* Program code: assign and use vectors */ \\
\\
glbFreeParams(vector1); \\
/* ... more vectors destroyed ... */ \\
\\
/* ... end of program or glbClearExperimentList ... */ 
}
\end{quote}

\section{Computing the simulated data}

Compared to existing experiments, which use real data, future experiments
uses simulated data. Thus, the {\em true parameter values} and their results in form of the reference event rate vectors are simulated. After setting the true parameter values, the {\em fit parameter values} can be varied in order to obtain information on the measurement performance for the given set of true parameter values. Therefore, it is often useful to show the results of a future measurement as function of the true parameter values for which the reference rate vectors are computed -- at least within the currently allowed ranges. The true parameter values for the vacuum neutrino oscillation parameters have to be set by the functions \GLB{glbSetOscillationParameters} and \GLB{glbSetRates} {\em before} any evaluation function is used and {\em after} the experiments have been initialized and the experiment parameters have been adjusted which could change the rates (such as baseline or target mass). This means that
after any change of an experiment parameter, {\tt glbSetRates} has to be called. Matter effects are automatically included as specified in the experiment definition. We have the following functions to assign and
read out the vacuum oscillation parameters:
\begin{function}
\GLBNS{glbSetOscillationParameters}
{\tt int glbSetOscillationParameters(const glb\_params in)} sets the
vacuum oscillation parameters to the ones in the vector {\tt in}.
\end{function}
\begin{function}
\GLBNS{glbGetOscillationParameters}
{\tt int glbGetOscillationParameters(glb\_params out)} returns the
vacuum oscillation parameters in the vector {\tt out}. The result of
the function is $0$ if the call was successful.
\end{function}
The reference rate vector is then computed with:
\begin{function}
\GLBNS{glbSetRates}
{\tt void glbSetRates()} computes the reference rate vector for the neutrino oscillation parameters set by {\tt glbSetOscillationParameters}. 
\end{function}
A complete example for a minimal \GLOBES\ program can be found on
Page~\pageref{ex:c}.

@@

\index{Systematics}
Calculating a $\chi^2$-value with or without systematics, but no correlations and degeneracies, is the simplest and fastest possibility to obtain high-level information on an experiment. In general, \GLOBES\ uses the six independent oscillation parameters $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, $\deltacp$, $\sdm$, $\ldm$, as well as the matter density 
scaling factor $\hat{\rho}$ of each experiment.\index{Matter density
scaling factor} Thus, there are six plus the number of experiments parameters determining the rate vectors. Using the matter density scaling factors  in addition to the oscillation parameters will allow the simulation of the correlations with matter density uncertainties: In this
approach, the matter density profile normalization $\hat{\rho}$ can 
be treated
as parameter to be measured by the experiment, where an external precision
given by observations is imposed (typically up to $5\%$). 
 For this section, it is important to keep in mind
that there are more parameters than just the oscillation parameters
determining the simple $\chi^2$. However, as we have described in the
last section, the mechanism for the matter density scaling factors
is hidden in the defintion of \GLB{glb\_params}: Each of the scaling
factors is initially set to $1.0$. Therefore, for the calculation of
$\chi^2$ with systematics only, we do not have to care about the
matter density scaling factors.

\example{Correlation between $\stheta$ and $\deltacp$}{
\label{ex:corrth13dcp}
\index{Two-parameter correlation}
A typical and fast application for \GLB{glbChiSys} is the visualization of
 two-parameter correlations using systematics only. For example, to
  calculate the two-parameter 
 correlation between $\stheta$ and $\deltacp$ at a neutrino factory, one
 can use the following code excerpt from {\tt example1.c}:
\begin{quote}
{\tt {\footnotesize
  /* Initialize parameter vector(s) and compute simulated data */ \\
  glbDefineParams(true\_values,theta12,theta13,theta23,deltacp,sdm,ldm); \\
  glbDefineParams(test\_values,theta12,theta13,theta23,deltacp,sdm,ldm);  \\
  glbSetOscillationParameters(true\_values); glbSetRates(); \\
\\
  /* Iteration over all values to be computed */ \\
  double x,y,res; \\   
  for(x=-4.0;x<-2.0+0.01;x=x+2.0/50) \\
  for(y=0.0;y<200.0+0.01;y=y+200.0/50) \\
  \{ \\
\hspace*{0.5cm} /* Set parameters in vector of test values */ \\
\hspace*{0.5cm} glbSetOscParams(test\_values,asin(sqrt(pow(10,x)))/2,GLB\_THETA\_13); \\
\hspace*{0.5cm} glbSetOscParams(test\_values,y*M\_PI/180.0,GLB\_DELTA\_CP); \\
  \\
\hspace*{0.5cm} /* Compute Chi2 for all loaded experiments and all rules */ \\
\hspace*{0.5cm} res=glbChiSys(test\_values,GLB\_ALL,GLB\_ALL); \\
\hspace*{0.5cm} AddToOutput(x,y,res); \\
  \} \\
} }
\end{quote}

\vspace*{-0.4cm}

@@

Keeping all oscillation parameters and matter density scaling factors fixed,
 one can use the following functions to obtain the total $\chi^2$ of all 
 specified oscillation channels including systematics:
\begin{function} 
\GLBNS{glbChiSys}
{\tt double glbChiSys(const glb\_params in,int exp, int rule)} returns
the $\chi^2$ for the (fixed) oscillation parameters {\tt in}, the
experiment number {\tt exp}, and the rule number {\tt rule}. For all
experiments or rules, use \GLB{GLB\_ALL} as parameter value.
\end{function}
Note that the result of {\tt glbChiSys} for all experiments or rules
corresponds to the sum of all of the individual {\tt glbChiSys} calls. 
This equality will not hold for the minimizors in the next sections anymore. 
 An example how to use  {\tt glbChiSys} can be found on page~\pageref{ex:corrth13dcp}.

\index{Systematics}
The treatment of systematics in \GLOBES\ is performed by the so-call
{\em pull method} with the help of auxiliary systematics parameters.
They are taken completely uncorrelated among different rules, and treated with simple Gaussian statistics. In general, a rule is a set 
of signal and background event rates coming from different oscillation
channels, where the event rates of all rule contributions are added.
For more details of the rule concept, see \Part~\ref{part:2} of this manual,
and for the treatment of systematics, see \Sec~\ref{sec:rules}.

 One example for a systematics parameter the signal normalization error, \ie, an error to the overall normalization of the signal. For illustration, we assume that the signal event rate in the $i$th bin $s_i^0$ of one oscillation channel is altered by the overall normalization auxiliary parameter\index{Auxiliary parameter} of this channel, \ie , 
\be
 s_i = s_i(n_s) = s_i^0 \cdot (1 + n_s),
\ee
where $n_s$ is the signal normalization parameter. The total number of events in the $i$th bin $x_i$ also includes the background event rates $b_i$, \ie, $x_i = s_i + b_i$, which may have their own systematics parameters.
In order to implement an overall signal normalization error $\sigma_{n_s}$,  the $\chi^2$, which includes all event rates $x_i$ of all bins, is minimized over the auxiliary parameter $n_s$:
\be
 \hat{\chi^2} = \underset{n_s}{\mathrm{min}} \left(  \chi^2(n_s, \hdots) + \frac{(n_s)^2}{\sigma_{n_s}^2} \right).
\ee 
This minimization is done independently for all auxiliary parameters of the rule. The total $\chi^2$ for the considered experiment is finally obtained by repreating this procedure for all rules and adding their $\chi^2$-values. In general, the situation is more complicated because of the usage of many systematical errors. More details about systematics parameters and the definition of signal, background, and oscillation channels can be found in \Sec~\ref{sec:rules}, too.

The systematics minimization of an experiment can be easily switched on and off with \GLB{glbSwitchSystematics}, \ie, one can also compute the $\chi^2$ with statistics only. In addition, several options for 
systematics are available, such as only using total event rates without
spectral information. For details, we refer to \Chapt~\ref{chapt:running}.

\chapter[Calculating $\chi^2$-projections: how one can include correlations]{Calculating $\boldsymbol{\chi^2}$-projections: how one can include correlations}
\label{chapt:correlations}

\index{Multi-parameter correlation}
This chapter deals with the rather complicated issue of $n$-parameter correlations. It is one of the greatest strengths of this software 
to include the full $n$-parameter correlation in the high-dimensional parameter space with reasonable effort. Of course, calculating $\chi^2$-projections is somewhat more complicated than using systematics only. Therefore, we use a simple step by step introduction to the problem.

@@

\index{Projection of manifold}
In principle, the precision of an individual parameter measurement including correlations in the $\chi^2$-approach can be obtained as the projection of the $n$-dimensional fit manifold onto the respective axis. Similarly, one can project the fit manifold onto a plane, such as the $\stheta$-$\deltacp$-plane, if one wants to explicitely show the allowed
region in this plane with all the other parameter correlations included. 
In practice, this projection is very difficult: a grid-based method would need $(N_{\mathrm{grid}})^n$ function calls of {\tt glbChiSys} to calculate the precision including the full $n$-parameter correlation, where $N_{\mathrm{grid}}$ is the number of points in each direction of the lattice. For example, taking only $N_{\mathrm{grid}}=20$ and $n=7$ (six oscillation parameters and matter density) would mean more than one billion function calls of {\tt glbChiSys}. One can easily image that this is too much for any sophisticated application.

\index{Minimizer}
The solution to this problem is using a local $n$-dimensional (local) minimizer for the procection instead of a grid-based method, where we will
illustrate this minimization process later. It turns out
that such a minimizer can include a full $6$-parameter correlation with of the order of $1\, 000$ function calls of {\tt glbChiSys}. It is a standard method which can be found in every good book for numerical calculation routines. Thus, for each point on the projection axis/plane, one can obtain a result within about $10$ to $30$ seconds on a modern computer, which means that the complete measurement precision for one fixed true parameter set can be obtained in as much as $10$ to $15$ minutes. One can easily imagine that such a minimizer makes more sophisticated applications possible with the help of overnight calculations, such as showing the dependencies on the true parameter values.

This approach also has one major disadvantage: One can not simply program a robust grid-based code and let it run, since using a local minimizer always means that one may end up in an unwanted local minimum and not in the investigated global one. Thus, one has to use some (analytical or numerical) knowledge on the topology of the fit manifold and start the local minimizer close enough to the investigated solution. Fortunately, this can be done quite straightfoward in most cases, since the structure of the neutrino oscillation formulas does not cause very complicated topologies of the fit manifolds. Especially the simulation of reactor experiments and conventional beams or superbeams is rather simple with purely numerical
approaches. Neutrino factories have, especially for small values of
$\theta_{13}$, a much more complicated topology. In this case, results
of the many analytical discussions of this issue can be used. This means
 that one can implicitely use the analytical knowledge to obtain better predictions for the measurement performances, since the minimizer has
 to be started only close enough to the exact solution by an ``educated
 guess''. One can easily imagine that the used methods then also depend
 on the region of the parameter space. In this manual, we only use
 examples with a neutrino factory, since some of these complications
 can be illustrated there.
